---
title: Making colorblind-friendly figures
date: '2023-03-07 11:00:00 -0800'
categories: [r, data]
tags: [r, data-structures, wrangling] # tags always lowercase
author: stefano
math: true # to enable math (not on by default)
output:
  html_document:
    keep_md: TRUE
---

\usepackage{amsmath}

There are many different data structure types in `R`, each with varying levels of complexity and uses. The simplest data structure in `R` is a vector. Vectors are one-dimensional arrays (i.e., sets of values) of a single class (such as characters, numbers, dates, etc.), and they have a similar structure to the mathematical concept of vectors. For example, the vector $\vec v$ with values 1, 4, 6, 2 would be:

$$\vec v = \begin{bmatrix}1 \\ 4 \\ 6 \\ 2\end{bmatrix}.$$

In `R` we can create vectors using the `c()` function, as follows:

```{r}
v <- c(1, 4, 6, 2)
v
```

(`R` prints vectors in a line rather than as columns to improve readability.)

Since vectors can only contain elements of a common type, `c()` will force all elements to be of a single class. In the example below, I create a vector with the number 1, the letter "a", today's date, and the value `TRUE` (a boolean value), but `c()` coerces all values to be characters:

```{r}
c(1, 'a', Sys.Date(), TRUE)
```

If we place two vectors (of the same class) side by side, we can create a matrix. Matrices are thus two-dimensional arrays 


```{r}
# 2D arrays: matrices
matrix(data = 1:8, nrow = 2)
matrix(data = 1:8, ncol = 2)
matrix(data = 1:8, ncol = 2, byrow = TRUE)

# matrix operations
matrix(1:9, ncol = 3)
matrix(1:9, ncol = 3) + 3
matrix(1:9, ncol = 3) * 2 # NOT matrix multiplication
matrix(1:9, ncol = 3) %*% 6:8 # matrix multiplication
matrix(1:9, ncol = 3) %*% matrix(1:6, ncol = 2)

# grouping objects with different and types: lists
l <- list(letters = c('a', 'b', 'c'),
          numbers = 1:10,
          today = Sys.Date())
l
l$letters
l$today

# lists with a table-like structure: data frames
#' note: `data.frame()` recycles elements if the longest vector is a multiple
data.frame(numbers = 1:10,
           letters = c('a', 'b', 'c'),
           today = Sys.Date())

data.frame(numbers = 1:10,
           letters = c('a', 'b'),
           today = Sys.Date())
```
