---
title: Basic Model Fitting in R
date: '2023-08-01 09:00:00 -0800'
categories: [R, R_Modelling]
tags: [R, modelling] # tags always lowercase
author: madison
output: 
  html_document:
    keep_md: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In previous categories we have looked at cleaning, summarizing, and subsetting data, with some minor calculations, but we haven’t yet looked at analyzing our data.

R is a very powerful tool for data analysis. We can fit linear models and view graphs. First, we will look at some basic data analysis processes in R.

Let’s re-load in our Gapminder data:

```{r}
link <- 'https://raw.githubusercontent.com/jstaf/gapminder/master/gapminder/gapminder.csv'
df <- read.csv(url(link))
head(df)
```

# Linear Modelling

Let’s say we want to fit a linear model to see if there is a relationship between population and time. First, let’s do this with the entire dataset with all countries. 

We know that there is Simple Linear Regression (SLR) and Multiple Linear Regression (MLR). See the pseudocode below for a layout on how to form both in R.

```{r}
# Simple Linear Regression
# model_name <- lm(Y ~ X, data = dataframe)

# Multiple Linear Regression
# model_name <- lm(Y ~ X1 + X2 + ..., data = dataframe)
# model_name <- lm(Y ~ ., data = dataframe) to use all variables except Y
```

Now that we have the layout of the code down, let's use Y as population and X as year for an SLR model.

```{r}
model1 <- lm(pop ~ year, data = df)
summary(model1)
```
Now, let's perform MLR with two variables, year and lifeExp.

```{r}
model2 <- lm(pop ~ year + lifeExp, data = df)
summary(model2)
```

Finally, let's perform MLR with all variables in the dataset.

```{r}
model3 <- lm(pop ~ ., data = df)
summary(model3)
```


Note that the outputs give us a lot of information. We see the default hypotheses tested, with p-values for each coefficient and for the model as a whole. We see the RSE and the degrees of freedom, F-test, R-squared values, and much more! There are a lot of insights that we can extract from this simple output.
